hbm_paper_results

### 2024-02-12 eda/lifespan_eda.Rmd
mostly plots
parsing lbcc harmonization

nrow(cn.df.unique) == n_unique(cn.df.unique$participant)
```

@@

```{r}
fwrite(cn.df.site, file=paste0(base_path, "/data/lifespan_CN_site-level.csv"))
```

>>>>

nrow(cn.df.unique) == n_unique(cn.df.unique$participant)

cn.df.unique%>%
  dplyr::filter(study != "UKB" & age_days <= (365.25*18)) %>%
  nrow()
```

@@

```{r}
fwrite(cn.df.site, file=paste0(base_path, "/data/lifespan_CN_site-level.csv"))
```
write out imagen data for testing
```{r}
imagen.df <- cn.df.site %>%
  dplyr::filter(study=="IMAGEN")
skim(imagen.df)

fwrite(imagen.df, file=paste0(base_path, "/data/imagen_CN.csv"))
```

### 2024-02-12 paper_lbcc_results.Rmd
parsing lbcc harmonization

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(data.table, ggplot2, tidyverse, ggseg, ggsegExtra, paletteer, ggpubr, gridExtra, lemon, parallel, rstatix, lme4, interactions, ggrepel, ggrain, skimr)
#Sys.setenv("MC_CORES"=10L)
options("mc.cores")

@@

#PHENO LISTS
pheno_list <- readRDS(file="cubic_scripts/R_scripts/pheno_list.rds")
vol_list_global <- readRDS(file="cubic_scripts/R_scripts/vol_list_global.rds")
vol_list_regions <- readRDS(file="cubic_scripts/R_scripts/Vol_list_regions.rds")
sa_list <- readRDS(file="cubic_scripts/R_scripts/SA_list.rds")
ct_list <- readRDS(file="cubic_scripts/R_scripts/CT_list.rds")

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(data.table, ggplot2, tidyverse, ggseg, ggsegExtra, paletteer, ggpubr, gridExtra, lemon, parallel, rstatix, lme4, interactions, ggrepel, ggrain, skimr, gam)
#Sys.setenv("MC_CORES"=10L)
options("mc.cores")

@@

#PHENO LISTS
pheno_list <- readRDS(file="cubic_scripts/R_scripts/pheno_list.rds")
pheno_list.z <- paste0(pheno_list, ".z")
vol_list_global <- readRDS(file="cubic_scripts/R_scripts/vol_list_global.rds")
vol_list_regions <- readRDS(file="cubic_scripts/R_scripts/Vol_list_regions.rds")
sa_list <- readRDS(file="cubic_scripts/R_scripts/SA_list.rds")
ct_list <- readRDS(file="cubic_scripts/R_scripts/CT_list.rds")

@@

#Site Effects

Test to see if centile score is associated with site in either ComBatLS or ComBat-GAM harmonized data.

## ComBat-GAM

```{r}
#clean predictions
gam_centile_df.clean <- gam_centile_df %>%
  mutate(study=sub("\\_.*", "", site),
         age_years=age_days/365.25) %>%
  dplyr::select(!contains(".z") & !ends_with("_total") & !contains("TBV")) #drop z-scores and non-centiles

skim(gam_centile_df.clean)
```

w/in feature ANOVAs

```{r}
gam.feat_to_test <- intersect(pheno_list, names(gam_centile_df.clean))

df_gam_feat_aov <- data.frame()
for(pheno in gam.feat_to_test){
  # print(pheno)
  df <- tidy(aov(gam_centile_df.clean[[pheno]] ~ gam_centile_df.clean[["study"]]))
  df$pheno <- pheno
  df_gam_feat_aov <- rbind(df_gam_feat_aov, df)
}

df_gam_feat_aov.study <- df_gam_feat_aov %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(gam.feat_to_test)), #apply fdr correction across all phenos tested
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
df_gam_feat_aov.study %>%
  ggplot() +
  geom_bar(aes(y=sig_fdr))
```

```{r}
gam_centile_df.clean %>%
  ggplot() +
  geom_boxplot(aes(x=study, y=WMV)) +
  theme(axis.text.x = element_text(angle=90, hjust=0.5))

summary(lm(WMV ~ study + age_days + sex, data = gam_centile_df.clean))
```

average w/in subject

```{r}
gam_centile_df.mean <- gam_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(participant) %>%
  summarize(study = unique(study),
            site = unique(site),
            age_days=unique(age_days),
            sex=unique(sex),
            mean_cent=mean(centile))
nrow(gam_centile_df.mean) == n_unique(gam_centile_df.mean$participant)
```

```{r}
gam_centile_df.mean %>%
  ggplot() +
  geom_boxplot(aes(x=study, y=mean_cent)) +
  theme(axis.text.x = element_text(angle=90, hjust=0.5))

gam_centile_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_cent, color=site), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_cent), method="lm") +
  theme(legend.position = "none") +
  facet_wrap(~study)
```

linear model

```{r}
summary(lm(mean_cent ~ study, data=gam_centile_df.mean))

summary(aov(mean_cent ~ study, data=gam_centile_df.mean))
pairwise.t.test(gam_centile_df.mean$mean_cent, gam_centile_df.mean$study, p.adjust.method = "fdr")

mean_cent_gam.lm <- lm(mean_cent ~ study + age_days + sex, data=gam_centile_df.mean)

gtsummary::tbl_regression(mean_cent_gam.lm) %>%
  gtsummary::add_global_p()
```

###Z-Scores

```{r}
#clean predictions
gam_z_df.clean <- gam_centile_df %>%
  mutate(study=sub("\\_.*", "", site),
         age_years=age_days/365.25) %>%
  dplyr::select(!any_of(pheno_list) & !ends_with("_total") & !contains("TBV")) #drop z-scores and non-centiles
```

w/in feature ANOVAs

```{r}
gam.feat_to_test.z <- paste0(gam.feat_to_test, ".z")
df_gam_feat_aov.z <- data.frame()
for(pheno in gam.feat_to_test.z){
  # print(pheno)
  df <- tidy(aov(gam_z_df.clean[[pheno]] ~ gam_z_df.clean[["study"]]))
  df$pheno <- pheno
  df_gam_feat_aov.z <- rbind(df_gam_feat_aov.z, df)
}

df_gam_feat_aov.z.study <- df_gam_feat_aov.z %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(gam.feat_to_test.z)), #apply fdr correction across all phenos tested
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
df_gam_feat_aov.z.study %>%
  ggplot() +
  geom_bar(aes(y=sig_fdr))
```

average w/in subject

```{r}
gam_z_df.mean <- gam_z_df.clean %>%
  pivot_longer(cols=any_of(pheno_list.z), names_to="pheno", values_to = "z_score") %>%
  group_by(participant) %>%
  summarize(study = unique(study),
            age_days=unique(age_days),
            sex=unique(sex),
            mean_z=mean(z_score))
nrow(gam_z_df.mean) == n_unique(gam_z_df.clean$participant)
```

```{r}
gam_z_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_z), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_z), method="lm") 
# +
#   facet_wrap(~study)
```

linear model

```{r}
summary(lm(mean_z ~ study, data=gam_z_df.mean))

summary(aov(mean_z ~ study, data=gam_z_df.mean))
pairwise.t.test(gam_z_df.mean$mean_z, gam_z_df.mean$study, p.adjust.method = "fdr")

mean_z_gam.lm <- lm(mean_z ~ study + age_days + sex, data=gam_z_df.mean)
summary(mean_z_gam.lm)
```

##ComBatLS

```{r}
#clean predictions
gamlss_centile_df.clean <- gamlss_centile_df %>%
  mutate(study=sub("\\_.*", "", site),
         age_years=age_days/365.25) %>%
  dplyr::select(!contains(".z") & !ends_with("_total") & !contains("TBV")) #drop z-scores and non-centiles
```

```{r}
gamlss_centile_df.clean %>%
  ggplot() +
  geom_boxplot(aes(x=study, y=WMV))

wmv_cent_gamlss.lm <- lm(WMV ~ study + age_days + sex, data=gamlss_centile_df.clean)

# gtsummary::tbl_regression(wmv_cent_gamlss.lm) %>%
#   gtsummary::add_global_p()

gamlss_centile_df.clean %>%
  ggplot() +
  geom_boxplot(aes(x=site, y=WMV, fill=study)) +
  theme(axis.text.x = element_text(angle=90), legend.position = "none")
```

w/in feature ANOVAs

```{r}
feat_to_test <- intersect(pheno_list, names(gamlss_centile_df.clean))

df_gamlss_feat_aov <- data.frame()
for(pheno in feat_to_test){
  # print(pheno)
  df <- tidy(aov(gamlss_centile_df.clean[[pheno]] ~ gamlss_centile_df.clean[["study"]]))
  df$pheno <- pheno
  df_gamlss_feat_aov <- rbind(df_gamlss_feat_aov, df)
}

df_gamlss_feat_aov.study <- df_gamlss_feat_aov %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(feat_to_test)), #apply fdr correction across all phenos tested
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
df_gamlss_feat_aov.study %>%
  ggplot() +
  geom_bar(aes(y=sig_fdr))

df_gamlss_feat_aov.study %>%
  dplyr::filter(sig_fdr==TRUE)
```
average w/in subject

```{r}
gamlss_centile_df.mean <- gamlss_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(participant) %>%
  summarize(study = unique(study),
            site = unique(site),
            age_days=unique(age_days),
            sex=unique(sex),
            log_age=unique(log_age),
            sexMale=unique(sexMale),
            fs_version=unique(fs_version),
            mean_cent=mean(centile))
nrow(gamlss_centile_df.mean) == n_unique(gamlss_centile_df.mean$participant)
```

```{r}
gamlss_centile_df.mean %>%
  ggplot() +
  geom_boxplot(aes(x=reorder(study, age_days), y=mean_cent)) + 
  theme(axis.text.x = element_text(angle=90))

gamlss_centile_df.mean %>%
  ggplot() +
  geom_violin(aes(x=reorder(study, age_days), y=mean_cent), scale="count") + 
  theme(axis.text.x = element_text(angle=90))

gamlss_centile_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_cent, color=site), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_cent), method="lm") +
  theme(legend.position = "none") +
  facet_wrap(~study)

gamlss_centile_df.mean %>%
  ggplot() +
  geom_boxplot(aes(x=site, y=mean_cent, fill=study)) + 
  theme(axis.text.x = element_text(angle=90), legend.position = "none")
```

linear model

```{r}
summary(lm(mean_cent ~ study, data=gamlss_centile_df.mean))

summary(aov(mean_cent ~ study, data=gamlss_centile_df.mean))
pairwise.t.test(gamlss_centile_df.mean$mean_cent, gamlss_centile_df.mean$study, p.adjust.method = "fdr")

mean_cent_gamlss.lm <- lm(mean_cent ~ study + age_days + sex + fs_version, data=gamlss_centile_df.mean)
summary(mean_cent_gamlss.lm)

mean_cent_gamlss.gam <- gam(mean_cent ~ study + s(log_age) + sex, data=gamlss_centile_df.mean)
summary(mean_cent_gamlss.gam)

mean_cent_gam.gam <- gam(mean_cent ~ study + s(log_age) + sex, data=gam_centile_df.mean)
summary(mean_cent_gam.gam)

gtsummary::tbl_regression(mean_cent_gamlss.lm) %>%
  gtsummary::add_global_p()
```

###Z-Scores

```{r}
#clean predictions
gamlss_z_df.clean <- gamlss_centile_df %>%
  mutate(study=sub("\\_.*", "", site),
         age_years=age_days/365.25) %>%
  dplyr::select(!any_of(pheno_list) & !ends_with("_total") & !contains("TBV")) #drop z-scores and non-centiles
```

w/in feature ANOVAs

```{r}
gamlss.feat_to_test.z <- paste0(feat_to_test, ".z")
df_gamlss_feat_aov.z <- data.frame()
for(pheno in gamlss.feat_to_test.z){
  # print(pheno)
  df <- tidy(aov(gamlss_z_df.clean[[pheno]] ~ gamlss_z_df.clean[["study"]]))
  df$pheno <- pheno
  df_gamlss_feat_aov.z <- rbind(df_gamlss_feat_aov.z, df)
}

df_gamlss_feat_aov.z.study <- df_gamlss_feat_aov.z %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(gamlss.feat_to_test.z)), #apply fdr correction across all phenos tested
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
df_gamlss_feat_aov.z.study %>%
  ggplot() +
  geom_bar(aes(y=sig_fdr))
```

average w/in subject

```{r}
gamlss_z_df.mean <- gamlss_z_df.clean %>%
  pivot_longer(cols=any_of(pheno_list.z), names_to="pheno", values_to = "z_score") %>%
  group_by(participant) %>%
  summarize(study = unique(study),
            age_days=unique(age_days),
            sex=unique(sex),
            mean_z=mean(z_score))
nrow(gamlss_z_df.mean) == n_unique(gamlss_z_df.clean$participant)
```

```{r}
gamlss_z_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_z, color=study), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_z), method="lm") +
  facet_wrap(~study)
```

linear model

```{r}
summary(lm(mean_z ~ study, data=gamlss_z_df.mean))

summary(aov(mean_z ~ study, data=gamlss_z_df.mean))
pairwise.t.test(gamlss_z_df.mean$mean_z, gamlss_z_df.mean$study, p.adjust.method = "fdr")

mean_cent_gamlss.lm <- lm(mean_z ~ study + age_days + sex, data=gamlss_z_df.mean)
summary(mean_cent_gamlss.lm)
```
